# Extending ggplot2

This is a personal summary of how to create a geom in `ggplot2` based on the vignette *Extending ggplot2* by Hadley Wickham and *Master Software Development in R* by Roger D. Peng, Seon Kross, and Brooke Anderson.

## Motivation

You may want to create your own geoms for two reasons:

* **Implement a new feature**: Maybe there is a feature that is not yet implemented.
* **Simplify a complex workflow**: With certain parts of analysis you may find yourself doing the same kind of plot repeatedly.

## Creating a new stat

To create a new geom or stat, you will need to follow the next steps:

  1.- Create a new ggproto that inherits from `Stat`/`Geom` and override the methods described below.

  2.- Write a layer function beginning with `stat_`/`geom_` with the following structure: default arguments, a `layer()` function, and `...` that will be used by the `params` argument.  The parameters given in `...` will either be arguments for the geom (if it is stat wrapper), arguments for the stat (if it is a geom wrapper), or asthetics to be set.

Here is a little example:

```{r}
library(ggplot2)
#------------------------------------------------------- 1. CREATE A NEW GGPROTO
StatChull <- ggproto(`_class`   = "StatChull",
                     #  Remember that we have have to inherit from Stat/Geom.
                     `_inherit` = Stat,
                     #  Does the computation
                     compute_group = function(data, scales) {
                     #  chull returns the elements that lie on the convex hull.
                     data[chull(data$x, data$y), , drop = FALSE]
                       },
                     #  Establishes the asthetics that must be present so as to
                     #  work.
                     required_aes = c("x", "y")
                     )
#----------------------------------------------------- 2. WRITE A LAYER FUNCTION
stat_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
                       position = "identity", na.rm = FALSE, show.legend = NA,
                       inherit.aes = TRUE, ...) {
  # Assures that the given parameters are stored correctly.
  layer(
    stat = StatChull, data = data, mapping = mapping, geom = geom,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}
# We can change the color of the convex hull for each group.
ggplot(mpg, aes(displ, hwy, colour = drv)) +
  geom_point() +
  stat_chull(fill = NA)
# Overrides the default geom.
ggplot(mpg, aes(displ, hwy)) +
  stat_chull(geom = "point", size = 4, colour = "red") +
  geom_point()
```

**N.B.**:  In case you are creating a package, you will either need to call `ggplot2::layer()` explicitly, or import the `layer()` function into your package namespace.

The subsequent example illustrates a stat that makes some tallies.

```{r}
#------------------------------------------------------- 1. CREATE A NEW GGPROTO
# This stat creates two points by which a regression line is defined.
StatLm <- ggproto("StatLm", Stat,
  required_aes = c("x", "y"),
  compute_group = function(data, scales) {
    rng <- range(data$x, na.rm = TRUE)
    grid <- data.frame(x = rng)
    mod <- lm(y ~ x, data = data)
    grid$y <- predict(mod, newdata = grid)
    grid
  }
)
#----------------------------------------------------- 2. WRITE A LAYER FUNCTION
stat_lm <- function(mapping = NULL, data = NULL, geom = "line",
                    position = "identity", na.rm = FALSE, show.legend = NA,
                    inherit.aes = TRUE, ...) {
  layer(
    stat = StatLm, data = data, mapping = mapping, geom = geom,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_lm()
```

In the coming example we have included parameters to the created stat.

```{r}
#------------------------------------------------------- 1. CREATE A NEW GGPROTO
StatLm <- ggproto("StatLm", Stat,
  required_aes = c("x", "y"),
  # We have added some parameters to the created stat that allows to change
  # formula by which the regression is computed.
  compute_group = function(data, scales, params, n = 100, formula = y ~ x) {
    rng <- range(data$x, na.rm = TRUE)
    grid <- data.frame(x = seq(rng[1], rng[2], length = n))
    mod <- lm(formula, data = data)
    grid$y <- predict(mod, newdata = grid)
    grid
  }
)
#----------------------------------------------------- 2. WRITE A LAYER FUNCTION
stat_lm <- function(mapping = NULL, data = NULL, geom = "line",
                    position = "identity", na.rm = FALSE, show.legend = NA,
                    inherit.aes = TRUE, n = 50, formula = y ~ x,
                    ...) {
  layer(
    stat = StatLm, data = data, mapping = mapping, geom = geom,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    # Includes the new parameters to the stat.
    params = list(n = n, formula = formula, na.rm = na.rm, ...)
  )
}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  stat_lm(formula = y ~ poly(x, 10)) +
  stat_lm(formula = y ~ poly(x, 10), geom = "point", colour = "red", n = 20)
```

The below example creates an automatic general parameter.

```{r}
#------------------------------------------------------- 1. CREATE A NEW GGPROTO
StatDensityCommon <- ggproto("StatDensityCommon", Stat,
  required_aes = "x",
  setup_params = function(data, params) {
    # If the bandwidth is specified, we do not have to perform its computation.
    if (!is.null(params$bandwidth))
      return(params)
    # We estimate the bandwidth for each group using the bw.nrd0 function; then,
    # we calculate the mean.
    xs <- split(data$x, data$group)
    bws <- vapply(xs, bw.nrd0, numeric(1))
    bw <- mean(bws)
    # Shows the calculated bandwidth.  This is a best practice.
    message("Picking bandwidth of ", signif(bw, 3))
    params$bandwidth <- bw
    params
  },
  compute_group = function(data, scales, bandwidth = 1) {
    d <- density(data$x, bw = bandwidth)
    data.frame(x = d$x, y = d$y)
  }
)
#----------------------------------------------------- 2. WRITE A LAYER FUNCTION
stat_density_common <- function(mapping = NULL, data = NULL, geom = "line",
                                position = "identity", na.rm = FALSE, show.legend = NA,
                                inherit.aes = TRUE, bandwidth = NULL,
                                ...) {
  layer(
    stat = StatDensityCommon, data = data, mapping = mapping, geom = geom,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    # We have to include the computed bandwidth to list of parameters.
    params = list(bandwidth = bandwidth, na.rm = na.rm, ...)
  )
}
# Since we have not indicated the bandwidth, the stat calculates it.
ggplot(mpg, aes(displ, colour = drv)) +
  stat_density_common()
# Once having specified the bandwidth parameter, the stats does not computes it.
ggplot(mpg, aes(displ, colour = drv)) +
  stat_density_common(bandwidth = 0.5)
```

We can return a calculated parameter as a default with the `default_aes` argument.

```{r}
# Since the layer has already defined, we only need to modify the ggproto
# object.
StatDensityCommon <- ggproto("StatDensity2", Stat,
  required_aes = "x",
  # Maps the density to y.
  default_aes = aes(y = ..density..),
  compute_group = function(data, scales, bandwidth = 1) {
    d <- density(data$x, bw = bandwidth)
    data.frame(x = d$x, density = d$y)
  }
)
# We use the calculated density by the created stat.
ggplot(mpg, aes(displ, drv, colour = ..density..)) +
  stat_density_common(bandwidth = 1, geom = "point")
```

## Creating a new geom

It is necessary to know about how it grid works so as to create a new geom.

The creation of a geom is similar to creating a stat. You always need to provide the following four elements to the ggproto object:

  1.- `required_aes` is a character vector of the required aesthetics.

  2.- `default_aes` sets up the default aesthetic values for certain arguments.

  3.- `draw_key` establishes the function used to draw the key in the legend (please see `?draw_key`).

  4.- `draw_panel()` Is a function that **returns a grid grob** and has the following three arguments (it is called once for **each panel**):

  a)  `data`: a data frame with a column **for each specified aesthetic**.

  b)  `panel_scales`: a list with information about the x and y scales **for the current panel**.

  c)  `coord`: an object that describes the coordinate system used by your plot.
  
  The last two parameters are not of much utility, except that they transform the data so as to be plotted.

```{r}
#------------------------------------------------------- 1. CREATE A NEW GGPROTO
GeomSimplePoint <- ggproto(`_class`   = "GeomSimplePoint",
                           `_inherit` = Geom,
  required_aes = c("x", "y"),
  default_aes = aes(shape = 19, colour = "black"),
  draw_key = draw_key_point,
  draw_panel = function(data, panel_scales, coord) {
    # Creates a data frame where position variables are scaled to the range (0, 1).
    coords <- coord$transform(data, panel_scales)
    # With the scaled data, you create the grid grob object.
    grid::pointsGrob(
      x = coords$x, y = coords$y,
      pch = coords$shape,
      gp = grid::gpar(col = coords$colour)
    )
  }
)
#----------------------------------------------------- 2. WRITE A LAYER FUNCTION
geom_simple_point <- function(mapping = NULL, data = NULL, stat = "identity",
                              position = "identity", na.rm = FALSE, show.legend = NA,
                              inherit.aes = TRUE, ...) {
  layer(
    geom = GeomSimplePoint, mapping = mapping,  data = data, stat = stat,
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}
# Uses the created geom.
ggplot(mpg, aes(displ, hwy)) +
  geom_simple_point()
```

